---
title: "Plot Functions Showcase"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Plot Functions Showcase}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 6,
  dpi = 150
)
```

This document showcases the plot functions in surveyzones to help
evaluate their usefulness.

## Setup: build a plan from example data

```{r build-plan, message=FALSE}
library(surveyzones)

tracts_sf <- surveyzones_example_tracts(n_tracts = 30, n_partitions = 2)

sparse_dists <- surveyzones_compute_sparse_distances(
  tracts_sf,
  engine = surveyzones_engine_haversine()
)

plan <- surveyzones_build_zones(
  sparse_distances = sparse_dists,
  tracts = sf::st_drop_geometry(tracts_sf),
  D_max = 10,
  max_workload_per_zone = 5,
  enforce_partition = TRUE
)
```

---

## 1. `surveyzones_plot_statistics()` -- Zone summary bar charts

Horizontal bar charts of zone-level metrics. Four modes: `"workload"`,
`"diameter"`, `"size"`, and `"all"`.

### Workload distribution

```{r stats-workload}
surveyzones_plot_statistics(plan, type = "workload")
```

### Diameter distribution

```{r stats-diameter}
surveyzones_plot_statistics(plan, type = "diameter")
```

### Number of tracts per zone

```{r stats-size}
surveyzones_plot_statistics(plan, type = "size")
```

### All metrics (faceted)

```{r stats-all}
surveyzones_plot_statistics(plan, type = "all")
```

---

## 2. `surveyzones_plot_connectivity()` -- Distance graph

Shows the connectivity graph: edges between tracts that are within `D_max`
of each other. Useful for understanding the spatial structure before
or after zoning.

```{r plot-connectivity}
surveyzones_plot_connectivity(tracts_sf, sparse_dists, D_max = 10)
```

### Tighter threshold

```{r plot-connectivity-tight}
surveyzones_plot_connectivity(tracts_sf, sparse_dists, D_max = 5)
```
