---
title: "Workload-Bounded Survey Zoning via Capacitated Facility Location with Geographic Diameter Bounds"
author:
  - name: Eduardo Leoni
    affiliation: IBGE (Instituto Brasileiro de Geografia e Estatística)
date: "2026-03-10"
format:
  html:
    number-sections: true
    toc: true
    toc-depth: 3
  pdf:
    number-sections: true
    toc: true
    documentclass: article
    geometry: margin=2.5cm
    fontsize: 11pt
execute:
  warning: false
  message: false
bibliography: references.bib
abstract: |
  Survey organizations must partition geographic regions into field zones that
  balance team workloads, remain geographically compact, and cover all sampled
  units. Standard facility location models---the p-median problem, the Warehouse
  Location Problem, and the Location Set Covering Problem---do not jointly
  address the combination of constraints specific to survey operations: workload
  capacity, a hard radius constraint on center-to-tract distance, and automatic
  determination of the number of zones. We present surveyzones, an integrated
  framework that enforces both constraints while searching for a feasible
  number of centers, using a two-phase decomposition. Phase 1 solves an
  uncapacitated p-median to find the minimum number of geographically feasible
  travel groups; Phase 2 splits oversized groups via capacitated sub-problems.
  The final zone count is an upper bound on the global minimum under both
  constraints. A three-level TSP-based sequencing algorithm then orders groups,
  zones, and tracts for efficient field traversal. We demonstrate the framework
  on two applications at contrasting scales: census tract zoning for a household
  survey in Rio de Janeiro (654 tracts, haversine distances) and biomarker
  collection logistics for the Brazilian National Health Survey across all 27
  state capitals (3,004 primary sampling units, OSRM road-network distances).
  The framework is implemented as an open-source R package.
---

```{r}
rdir <- '~/gitlab/ibgeba/pns.zonas/'

```
# Introduction

Survey organizations face a fundamental operational challenge: how to partition
geographic regions into zones for field teams such that each team has a
manageable, balanced workload; field operations remain efficient (teams do not
waste time traveling); and zones are geographically cohesive and traversable
within a reasonable time budget.

This challenge arises in every household survey that deploys interviewers or
collection teams across a territory. When a national health survey adds
biomarker collection---requiring cold-chain logistics and same-day sample
processing---the zoning problem becomes critical: teams must complete all
scheduled collections within a zone and return to the laboratory support point
before samples degrade. Inefficient zoning translates directly into fewer
collections per day, extended field periods, and higher costs.

Standard facility location models in the operations research
literature---such as the Location Set Covering Problem [LSCP; @toregas1971],
the p-median problem [@hakimi1964], and the Warehouse Location Problem
[WLP; @kuehn1963]---were designed for retail, emergency services, or supply
chain contexts. While capacitated variants with distance cutoffs exist in the
literature, no standard formulation combines workload capacity, a hard radius
constraint ($D_{\max}$), and automatic determination of $K$ within a
survey-oriented operational pipeline. Moreover, these models typically require
the number of facilities $K$ as an input parameter.

We present surveyzones, an integrated framework that enforces both
constraints while searching for a feasible $K$, through a two-phase
decomposition: an uncapacitated p-median finds the minimum number of
geographically feasible *travel groups*, which are then split by workload
capacity into final zones. The resulting three-level hierarchy (groups $\to$
zones $\to$ tracts) is exploited by a TSP-based sequencing algorithm for
efficient field traversal.

The remainder of this paper is organized as follows. @sec-constraints defines
the survey-specific constraints that motivate the model. @sec-related reviews
related facility location models. @sec-model presents the surveyzones
formulation, the two-phase decomposition, and the implementation pipeline.
@sec-applications demonstrates the framework on two applications at
contrasting scales. @sec-discussion discusses limitations and future
directions.


# Survey-Specific Constraints {#sec-constraints}

**Workload capacity.** Field teams have finite capacity. Zones must respect a
maximum workload per team ($W_{\max}$), and we assume $s_i \leq W_{\max}$ for
every tract $i$ (no single tract exceeds team capacity). Unlike logistics
applications where per-stop demand is unknown in advance, survey workload is
largely predetermined: the target number of households per tract is fixed by
the sampling frame. Expected service time per tract is therefore estimable
before deployment. Note that the objective function minimizes total travel
distance, not workload variance: "workload-bounded" refers to satisfying the
capacity constraint, not to equitable distribution of workload across zones.

**Geographic reach.** Teams cannot efficiently operate in zones that are
geographically dispersed. $D_{\max}$ defines the maximum travel time (or
distance) allowed between any tract and its zone center---a **radius
constraint**. Any tract more than $D_{\max}$ from the zone center is excluded
from that zone. The **diameter** of a zone, $\max_{i,j \in Z} d_{ij}$, is the
maximum pairwise distance between any two tracts in the zone. Since every
tract is within $D_{\max}$ of the zone center, the triangle inequality gives a
derived diameter bound of $2 \times D_{\max}$.^[This bound is exact under a
symmetric metric distance (e.g., haversine, or a symmetrized travel-time
matrix). With directed/asymmetric travel times or mixed engines (e.g., OSRM
for some pairs and haversine fallback for others), the metric assumptions may
not hold exactly, so $2 \times D_{\max}$ should be treated as an operational
rule of thumb rather than a formal guarantee. In practice, observed zone
diameters are often well below $2
\times D_{\max}$ because the sum-of-distances objective favors compact
assignments.]

**$K$ as output, not input.** The number of centers $K$ in the MILP is not
known a priori---it depends on the spatial distribution of tracts, the
$D_{\max}$ constraint, and the workload capacity. Existing models require $K$
as an input. A practical framework for survey operations must determine a
feasible $K$ automatically, rather than requiring the user to specify it.


# Related Work {#sec-related}

## p-Median

The p-median problem minimizes total weighted travel distance from demand
points to their nearest open facility, given $K$ facilities:

$$\min \sum_i \sum_j d_{ij} \cdot w_i \cdot x_{ij}$$

subject to $\sum_j y_j = K$, $\sum_j x_{ij} = 1\ \forall i$, and
$x_{ij} \leq y_j\ \forall i,j$.

The p-median minimizes total travel distance (demand-weighted by $w_i$);
surveyzones minimizes unweighted total travel time. Workload ($s_i$) enters
only through the capacity constraint (@eq-capacity), not the objective
(@eq-objective): the objective is pure distance minimization, while workload
constrains the feasible set. Both models minimize a sum-of-distances
objective, but p-median imposes no workload capacity and no zone-size bound.
It may produce zones with highly unequal workloads, since geographic
proximity---not workload---drives assignment. It also allows zones to span
arbitrarily large areas, which is operationally infeasible for survey teams.

## Warehouse Location Problem

The Warehouse Location Problem (WLP) minimizes fixed facility-opening costs
plus variable transportation costs, subject to warehouse capacity:

$$\min \sum_j f_j y_j + \sum_i \sum_j c_{ij} x_{ij}$$

subject to $\sum_j x_{ij} = 1\ \forall i$ and
$\sum_i s_i x_{ij} \leq S_j y_j\ \forall j$.

surveyzones is structurally a capacitated facility location problem: every
tract is a candidate zone center, opening costs are uniform, and the
assignment cost is travel time. Unlike WLP---where $K$ is determined by
trading off fixed opening costs against transport savings---surveyzones searches for a feasible $K$ via an external search loop, rather
than requiring $K$ as input. The critical addition absent from
standard WLP is the $D_{\max}$ constraint: WLP respects capacity but permits
zones of arbitrary geographic extent.

## Location Set Covering Problem

The Location Set Covering Problem (LSCP) minimizes the number of facilities
needed to ensure every demand point is within $D_{\max}$ of some open
facility:

$$\min \sum_j y_j \quad \text{s.t.} \quad \sum_{j:\, d_{ij} \leq D_{\max}} y_j \geq 1\ \forall i.$$

LSCP shares surveyzones' goal of using as few zones as possible, and it uses
$D_{\max}$ as a parameter. However, it produces no zone assignments---only
facility locations---and its objective contains no workload balancing or
distance minimization. Any two configurations achieving the same minimum $K$
are equally optimal, even if one requires twice the total travel time.

## Territory Design and Related Applications

The territory design literature [@rios-mercado2009; @kalcsics2005] addresses
geographic partitioning with workload balance and compactness objectives,
primarily in sales force [@zoltners1983] and political districting contexts.
@camacho2015 partition a city into police patrol sectors balancing workload,
compactness, and geographic constraints---a problem structurally analogous to
survey zoning. @begur1997 present an integrated spatial DSS for scheduling
and routing home-health-care nurses that combines territory design with
workload balancing for field workers. These models typically enforce
contiguity and balance multiple criteria (population, workload, revenue) but
do not impose a hard diameter bound. The number of territories is usually
fixed exogenously. surveyzones shares the workload balance objective but adds
$D_{\max}$ pre-filtering and a $K$-feasibility search that does not require
the user to specify the number of zones.

## Software

The `spopt` library in the PySAL ecosystem [@rey2021] provides Python
implementations of p-median, p-center, LSCP, MCLP, and capacitated variants.
It is the closest existing tool to surveyzones but lacks $D_{\max}$
pre-filtering and automatic $K$-search. The `ompr` R package
[@schumacher2020] provides the MILP modelling language used by surveyzones.

## Summary

| | LSCP | p-median | WLP | Territory design | **surveyzones** |
|---|:---:|:---:|:---:|:---:|:---:|
| Workload capacity | | | $\checkmark$ | $\checkmark$ | $\checkmark$ |
| Radius constraint ($D_{\max}$) | | | | | $\checkmark$ |
| 100% assignment | | $\checkmark$ | $\checkmark$ | $\checkmark$ | $\checkmark$ |
| Distance minimization | | $\checkmark$ | $\checkmark$ | $\checkmark$ | $\checkmark$ |
| $K$ not required as input | $\checkmark$ | | | | $\checkmark$ |

: Constraint coverage across facility location models. LSCP ensures coverage (reachability within $D_{\max}$) but does not assign demand points to facilities. {#tbl-comparison}


# The surveyzones Model {#sec-model}

## Formulation

For a fixed number of zones $K$, surveyzones solves a capacitated p-median
MILP:

$$\min \sum_i \sum_k d_{ik} \cdot z_{ik}$$ {#eq-objective}

subject to:

$$\sum_k z_{ik} = 1 \quad \forall i \quad \text{(all tracts assigned)}$$ {#eq-assignment}
$$\sum_k y_k = K \quad \text{(exactly } K \text{ zones)}$$ {#eq-zones}
$$z_{ik} \leq y_k \quad \forall i,k \quad \text{(assign only to open zones)}$$ {#eq-open}
$$\sum_i s_i \cdot z_{ik} \leq W_{\max} \cdot y_k \quad \forall k \quad \text{(workload capacity)}$$ {#eq-capacity}

where $d_{ik}$ is the travel time from tract $i$ to candidate center $k$,
$z_{ik} \in \{0,1\}$ indicates whether tract $i$ is assigned to zone $k$,
$y_k \in \{0,1\}$ indicates whether candidate $k$ is opened as a zone
center, and $s_i$ is the workload (expected service time) of tract $i$. The
candidate center set equals the full set of tracts: every tract may serve as a
zone center.

The geographic reach constraint ($d_{ik} \leq D_{\max}$) is enforced in
pre-processing: any $(i, k)$ pair with travel time exceeding $D_{\max}$ is
excluded from the model, setting $z_{ik} = 0$ by construction. Note that
$D_{\max}$ and $d_{ik}$ share the same units, determined by the distance
engine: minutes for OSRM travel times, kilometres for haversine distances.
This $D_{\max}$ pre-filtering reduces the number of decision variables
substantially and keeps zones geographically compact without adding
constraints to the MILP.


## Minimum-$K$ Search and Hierarchical Structure {#sec-k-discovery}

The operative sizing parameter is $W_{\max}$: it defines how many tracts (or
how much total service time) a team can handle and determines the actual size
of zones. However, $K$ need not be derived directly from workload. The default
strategy, `uncap_then_split`, decouples the $K$-search from workload balancing
by producing a **three-level hierarchy: travel groups $\to$ zones $\to$
tracts**.

Algorithm 1: uncap_then_split (outline; details in Phases 1--2 below)

1. Filter edges: retain only $(i, k)$ pairs with $d_{ik} \leq D_{\max}$
2. Identify connected components in the filtered graph
3. **Phase 1** --- For each component:
      for $K = 1, 2, \ldots$ until the uncapacitated MILP is feasible
      (i.e., all tracts assigned to a center within $D_{\max}$):
           solve uncapacitated p-median $\to$ **travel groups**
4. **Phase 2** --- For each travel group with $\sum_i s_i > W_{\max}$:
      set $K_{\text{split}} = \lceil \sum_i s_i / W_{\max} \rceil$;
      solve capacitated p-median; increment $K_{\text{split}}$ on infeasibility
      $\to$ **zones**
5. Return zones (groups within capacity become single-zone groups)


### Phase 1: Geographic grouping

Phase 1 performs a feasibility search over $K$ using the p-median assignment
structure with $D_{\max}$-filtered edges. Specifically, it solves the
*uncapacitated* p-median---the formulation above without constraint
(@eq-capacity)---starting from $K_0 = 1$ and incrementing until the assignment
MILP is feasible (all tracts can be assigned to centers within $D_{\max}$).
Starting from 1 rather than a workload-derived lower bound $\lceil n /
W_{\max} \rceil$ is deliberate: Phase 1 ignores workload entirely and seeks the
minimum $K$ for geographic feasibility alone, which may be much smaller. (When
the $D_{\max}$ graph contains $C$ disconnected components, $K \geq C$ is a
necessary condition for feasibility; the connectivity analysis from the
pipeline's Stage 2 already identifies these components, and each is solved
independently.)

$$K^*_{\text{uncap}} = \min \{ K \geq 1 : \text{uncapacitated MILP feasible} \}$$

The distance-minimization objective does not drive $K$; it selects the
best assignment at each $K$. Without capacity constraints the MILP has fewer
variables and constraints, and convergence is fast. Each resulting cluster
defines a **travel group**---a set of geographically cohesive tracts sharing
the same Phase 1 center.

### Phase 2: Workload splitting

Any travel group whose total workload exceeds $W_{\max}$ is split into
smaller **zones** by a single round of capacitated p-median restricted to that
group. Each sub-problem involves only the tracts within one group and is
solved instantly. The $D_{\max}$ radius constraint is re-applied in Phase 2:
the sub-problem uses the same $D_{\max}$-filtered distance matrix, so
sub-zone centers respect the same geographic reach as Phase 1. Groups already
within capacity become single-zone groups.
If the capacitated sub-problem at $K_{\text{split}} = \lceil W_{\text{group}}
/ W_{\max} \rceil$ is infeasible (e.g., no set of $K_{\text{split}}$ distinct
centers can partition the group within both $D_{\max}$ and $W_{\max}$), the
implementation retries with increasing $K$ up to $n$ (trivially feasible:
every tract becomes its own zone). If the solver still fails after all
attempts (only possible due to solver time limits, not mathematical
infeasibility), the implementation returns an infeasible status for that
partition rather than accepting an oversized zone.
The parent group relationship is preserved, so all sub-zones of a split group
share the same group identifier.

The `uncap_then_split` decomposition is a heuristic: because Phase 1 ignores
workload, the travel groups it produces may not align with the globally
optimal capacitated partition, and the final zone count after splitting is an
upper bound on the true minimum $K$ under both constraints. The decomposition
trades global optimality for computational tractability and the travel group
hierarchy. Crucially, this hierarchy is
not merely cosmetic: because every tract in a group is within $D_{\max}$ of
the group center, any two tracts in the same group are reachable within $2
\times D_{\max}$ of each other (under metric assumptions above). The groups
produced by Phase 1 are large geographically coherent clusters under
$D_{\max}$ because the uncapacitated solve prioritizes minimum center count
before workload splitting. Zones within a group are therefore
*operationally interchangeable*---field coordinators can reassign tracts
between zones within a group, adjust team schedules, or absorb workload
fluctuations without violating the geographic diameter bound. This
interchangeability is a direct consequence of the group-first priority:
groups are optimized for geographic cohesion before workload splits them
into smaller units.

The two-phase decomposition has several advantages over solving the full
capacitated problem directly. First, the uncapacitated Phase 1 converges
quickly because the MILP is smaller and the $K$-search only needs geographic
feasibility. Second, Phase 2 sub-problems are tiny (typically 2--10 tracts
each) and solved independently. Third, the resulting travel group structure
provides a natural hierarchy for downstream sequencing and for human-readable
zone identifiers.

### Alternative strategy: direct capacitated solve {#sec-direct}

An alternative `direct` strategy solves the full capacitated p-median MILP
directly, incrementing $K$ from a workload-derived lower bound $\lceil
W_{\text{total}} / W_{\max} \rceil$. This produces the minimum number of
zones under both constraints---potentially fewer than `uncap_then_split`---but
does not inherently produce a group hierarchy. Groups can be recovered
post-hoc via a lightweight set cover problem: given the fixed zones, find the
fewest group centers such that every zone's tracts are within $D_{\max}$ of
their group center. Since every zone center already covers its own zone (the
trivial solution $K_{\text{groups}} = K_{\text{zones}}$ is always feasible),
set cover minimizes over this upper bound.

The two strategies represent different lexicographic priorities:

- **`uncap_then_split`** (group-first): minimise groups, then split for
  capacity. Produces the fewest, largest groups at the cost of potentially
  more zones.
- **`direct`** (zone-first): minimise zones under capacity, then recover
  groups post-hoc.

The choice depends on operational priorities: when field coordination at the
group level dominates (e.g., team assignment, logistics planning), the
group-first decomposition is preferable; when minimising the total number of
field units is paramount, the direct strategy is appropriate.

This architecture combines a $K$-feasibility search (inspired by LSCP),
p-median's distance minimization, WLP's capacity constraints, and $D_{\max}$
pre-filtering. The user specifies $W_{\max}$ and $D_{\max}$; the framework
determines a feasible $K$ automatically.


## Implementation Pipeline {#sec-pipeline}

surveyzones operates in four stages.

**Stage 1---Distance computation.** Pairwise travel times between tract access
points are computed using a pluggable distance engine. The primary engine uses
OSRM [@luxen2011] for road-network travel times. Where OSRM cannot route a
pair (e.g., islands, river-only access), haversine great-circle distance is
used and scaled to a deliberately harsh travel time penalty (equivalent to
$\sim$1 km/h), discouraging cross-barrier assignments while keeping every
tract assignable as a last resort. Distances are stored as sparse
origin--destination pairs, which facilitates $D_{\max}$ filtering and
disk caching.

**Stage 2---Connectivity analysis.** The sparse distance table is filtered by
$D_{\max}$ to produce a connectivity graph: only tract--center pairs within
$D_{\max}$ are retained as feasible assignments. Disconnected components are
identified via graph analysis. Partition constraints (e.g., by state or by
agency jurisdiction) applied via `enforce_partition` further subdivide the
problem. Each component of each partition is passed to Stage 3 as an
independent sub-problem, enabling decomposition and potential parallelism
across partitions.

**Stage 3---Zone construction.** The two-phase `uncap_then_split` strategy
(@sec-k-discovery) is executed for each sub-problem. The MILP is formulated
using the `ompr` modelling language [@schumacher2020] and solved with HiGHS
[@huangfu2018] (or GLPK/CBC as alternatives) via the ROI interface. Results
are cached to disk: repeated calls with identical parameters return instantly
without re-running the solver.

**Stage 4---Sequencing.** Visit order is computed at three hierarchical
levels, exploiting the group $\to$ zone $\to$ tract structure:

1. **Group ordering** (within each partition): groups are ordered via TSP on
   the minimum pairwise tract distances between representative zones of each
   group.
2. **Zone ordering** (within each group): zones are ordered via TSP on the
   minimum pairwise tract distances between zones.
3. **Tract ordering** (within each zone): tracts are ordered via TSP. Tract
   sequences are oriented so that each zone's entry tract is closest to the
   previous zone's exit tract.

Missing cross-zone distances are filled using haversine with a speed penalty
of 0.1 km/h---more severe than the 1 km/h used in Stage 1. The distinction is
deliberate: during zone construction, the fallback must keep every tract
assignable (penalty large enough to discourage, but finite); during sequencing,
missing pairs represent true barriers between zones that should be visited last
in any route. At the end
of sequencing, zone IDs are renamed to a hierarchical format
`{partition}_{group}.{position}` (e.g., `33_1.001`, `33_1.002`), making the
travel group structure immediately visible in the zone identifier.


# Applications {#sec-applications}

The applications below are demonstrations of feasibility and operational
relevance rather than causal evaluation of algorithmic performance. We
illustrate the framework on two cases at contrasting scales: census tract
zoning for a simulated household survey in Rio de Janeiro (@sec-census)
and biomarker collection logistics for the Brazilian National Health Survey
across all 27 state capitals (@sec-pns).

## Census Tract Zoning: Rio de Janeiro {#sec-census}

### Data and setup

We construct a realistic survey scenario using publicly available data from
the 2022 Brazilian Census. Census tracts for the city of Rio de Janeiro
(municipality code 3304557) are obtained via the `censobr` package
[@pereira2024]. Tracts with zero population are excluded.

Rio de Janeiro has a single administrative district subdivided into 33
subdistricts. We use subdistricts as strata and draw a stratified random
sample of up to 20 tracts per subdistrict.

```{r}
#| label: census-setup
#| eval: false

library(censobr)
library(cnefetools)
library(surveyzones)
library(dplyr)
library(sf)

set.seed(20091975)

tracts_rj <- read_tracts(year = 2022, dataset = "Preliminares") |>
  collect() |>
  filter(code_muni == 3304557, V0001 > 0) |>
  mutate(subdistrict = substr(code_tract, 1, 11))

sampled_tracts <- tracts_rj |>
  slice_sample(n = 20, by = subdistrict)
```

We enrich the sample with favela status from census tract characteristics:
tracts belonging to recognised *aglomerados subnormais* (informal settlements)
are flagged. This defines a **partition**: zones must not mix *comunidade*
tracts with regular tracts, reflecting distinct access conditions and
interviewer protocols.

```{r}
#| label: census-partition
#| eval: false

tract_chars <- read_tracts(year = 2022, dataset = "ResponsavelRenda") |>
  collect() |>
  filter(code_muni == 3304557) |>
  select(code_tract, code_favela)

sampled_tracts <- sampled_tracts |>
  left_join(tract_chars, by = "code_tract") |>
  mutate(partition_id = if_else(!is.na(code_favela), "comunidade", "regular"))
```

### Representative access points

Geocoded household addresses from the CNEFE (Cadastro Nacional de Endereços
para Fins Estatísticos) are downloaded via the `cnefetools` package. For each
sampled tract, `surveyzones_representative_points()` computes the
highest-density location using kernel density estimation on the geocoded
addresses. These representative points approximate interviewer entry
locations rather than geometric centroids, providing more operationally
relevant origins for distance computation.

```{r}
#| label: census-access-points
#| eval: false

cnefe_rj <- read_cnefe(code_muni = 3304557) |> collect()

cnefe_sample <- cnefe_rj |>
  mutate(code_tract = substr(COD_SETOR, 1, 15)) |>
  filter(code_tract %in% sampled_tracts$code_tract)

access_pts <- surveyzones_representative_points(
  cnefe_sample,
  tract_col = "code_tract"
)
```

### Zone construction

Pairwise haversine distances are computed between access points. The
capacitated p-median is solved with $D_{\max} = 3$ km and $W_{\max} = 5$
tracts per zone. We build two plans---one ignoring the favela partition and
one enforcing it---to illustrate the effect of operational constraints.

```{r}
#| label: census-zones
#| eval: false

distances <- surveyzones_compute_sparse_distances(
  access_points = access_pts,
  engine = surveyzones_engine_haversine(units = "km")
)

tracts_df <- st_drop_geometry(access_pts) |>
  mutate(expected_service_time = 1L)

plan_no_partition <- surveyzones_build_zones(
  sparse_distances      = distances,
  tracts                = tracts_df,
  D_max                 = 3,
  max_workload_per_zone = 5L,
  enforce_partition      = FALSE,
  solver                = "highs"
)

plan_partition <- surveyzones_build_zones(
  sparse_distances      = distances,
  tracts                = tracts_df,
  D_max                 = 3,
  max_workload_per_zone = 5L,
  enforce_partition      = TRUE,
  solver                = "highs"
)
```

### Results

```{r}
#| label: census-results-code
#| eval: false

plan_no_partition <- surveyzones_sequence(plan_no_partition, distances)
plan_partition    <- surveyzones_sequence(plan_partition, distances)

surveyzones_compare(list(
  no_partition = plan_no_partition,
  partition    = plan_partition
))
```

```{r}
#| label: census-results
#| echo: false

census_cache <- readRDS("census_results.rds")
plan_no_partition <- census_cache$plan_no_partition
plan_partition    <- census_cache$plan_partition

comp <- census_cache$comparison
comp$plan_name <- c("No partition", "Partition")
names(comp) <- c("Plan", "Partitions", "Tracts", "Zones",
  "Wkld min", "Wkld max", "Wkld mean",
  "Diam min (km)", "Diam max (km)", "Diam mean (km)", "Obj (km)")
comp |>
  tinytable::tt(
    caption = "Census tract zoning: comparison of plans with and without favela partitioning.",
    digits = 2
  )
```

The stratified sample yields `r census_cache$n_tracts` tracts across
`r census_cache$n_subdistricts` subdistricts, of which
`r census_cache$n_comunidade` are classified as *comunidade* and
`r census_cache$n_regular` as regular. Zoning uses only the observed sampled
tracts; survey design weights do not affect the grouping.

Without partitioning, the algorithm generates
`r census_cache$comparison$n_zones[1]` zones (mean
`r round(census_cache$comparison$workload_mean[1], 1)` tracts/zone, max
diameter `r round(census_cache$comparison$diameter_max[1], 1)` km). Enforcing
the favela partition increases the count to
`r census_cache$comparison$n_zones[2]` zones (mean
`r round(census_cache$comparison$workload_mean[2], 1)` tracts/zone, max
diameter `r round(census_cache$comparison$diameter_max[2], 1)` km): isolated
*comunidade* tracts that would otherwise join nearby regular tracts now form
their own zones. Observed diameters remain well below the theoretical bound
$2 \times D_{\max} = 6$ km, because the sum-of-distances objective favors
compact assignments. The partitioned plan has a lower objective value (total
travel distance) because the additional zones provide more centers, reducing
average assignment distances. However, this comes at an operational cost:
an increase in the number of zones necessary to cover all the assignments.

The partition guarantee ensures that field teams assigned to
informal settlements operate exclusively within that context---a common
operational requirement reflecting distinct interviewer training, security
protocols, and access logistics.

```{r}
#| label: census-stats
#| echo: false
#| fig-width: 8
#| fig-height: 4

surveyzones::surveyzones_plot_statistics(plan_partition, type = "all")
```

@fig-census-maps shows the resulting zone layouts. Each tract is coloured by
travel group; convex hulls (with a 300 m buffer) delineate individual zones.
Without partitioning, zone hulls may span both *comunidade* and regular tracts;
with partitioning, each zone is confined to a single stratum.

```{r}
#| label: fig-census-maps
#| echo: false
#| fig-width: 10
#| fig-height: 5
#| fig-cap: "Rio de Janeiro census tract zoning. Left: no partition constraint. Right: favela/regular partition enforced. Tracts are coloured by travel group; translucent hulls delineate zones."

library(sf)
library(dplyr)

make_map_gg <- function(plan, title = NULL) {
  pts <- plan$access_points |>
    dplyr::left_join(plan$assignments, by = "tract_id", suffix = c("", "_plan")) |>
    sf::st_transform(4326L)

  # Color by group_id
  if (!is.null(plan$zone_sequence) && "group_id" %in% names(plan$zone_sequence)) {
    group_lookup <- plan$zone_sequence |> dplyr::select(zone_id, group_id)
    pts <- pts |>
      dplyr::left_join(group_lookup, by = "zone_id") |>
      dplyr::mutate(grp = factor(group_id))
  } else {
    pts <- pts |> dplyr::mutate(grp = factor(zone_id))
  }

  hulls <- pts |>
    sf::st_transform(5880L) |>
    dplyr::group_by(zone_id, grp) |>
    dplyr::summarise(
      geometry = sf::st_buffer(sf::st_convex_hull(sf::st_union(geometry)), dist = 300),
      .groups = "drop"
    ) |>
    sf::st_transform(4326L)

  ggplot2::ggplot() +
    ggplot2::geom_sf(data = hulls, ggplot2::aes(fill = grp),
                     alpha = 0.3, colour = "grey40", linewidth = 0.2) +
    ggplot2::geom_sf(data = pts, ggplot2::aes(colour = grp),
                     size = 0.6, show.legend = FALSE) +
    ggplot2::scale_fill_viridis_d(option = "turbo", guide = "none") +
    ggplot2::scale_colour_viridis_d(option = "turbo", guide = "none") +
    ggplot2::labs(title = title) +
    ggplot2::theme_void(base_size = 9) +
    ggplot2::theme(plot.title = ggplot2::element_text(hjust = 0.5, size = 10))
}

p1 <- make_map_gg(plan_no_partition, "No partition")
p2 <- make_map_gg(plan_partition, "With partition")
patchwork::wrap_plots(p1, p2, ncol = 2)
```


## National Health Survey: Biomarker Collection {#sec-pns}

### Operational context

The Brazilian National Health Survey (PNS) collects population health data
from a representative sample of households. For the 2026 cycle, biomarker
collection (blood and urine) is conducted in a subsample of households across
all 27 state capitals and their respective metropolitan regions, with a
completion deadline of October 2026.

The operational flow is:

1. Collection teams depart from a laboratory support point.
2. They travel to a **geographic zone** (a group of primary sampling
   units---UPAs).
3. They complete scheduled collections within that zone.
4. They return to the laboratory for sample processing.

Biological samples require immediate cold storage after collection. Zones must
therefore be compact enough for teams to complete all scheduled visits and
return within the cold-chain time window. The zoning problem is not merely an
efficiency concern---it directly constrains the feasibility of the collection
protocol.

### Parameters

Based on operational requirements, the following parameters were adopted:

| Parameter | Value | Description |
|-----------|-------|-------------|
| $D_{\max}$ | 10 min | Maximum OSRM travel time from any UPA to its zone center |
| $W_{\max}$ | 6 UPAs | Maximum UPAs per zone |
| Partition | By state (UF) | Each state solved independently |
| Solver | HiGHS | Relative tolerance 1%, time limit 180s per state |

### Distance computation

Distances were computed using OSRM road-network travel times as the primary
source (`surveyzones_engine_osrm(measure = "duration")`). For UPA pairs where
OSRM could not compute a route (islands, river-only access), haversine
distance scaled by a factor of 60 (equivalent to 1 km/h travel speed) was
used as fallback. The 60$\times$ multiplier was chosen so that a 1 km
haversine gap produces a 60-minute penalty---well above the 10-minute
$D_{\max}$---making cross-barrier assignments prohibitively expensive while
keeping every tract technically assignable as a last resort.

```{r}
#| label: pns-distances
#| eval: false

# Representative points per UPA (kernel density on geocoded addresses)
access_points <- surveyzones_representative_points(households, tract_col = "upa")

# OSRM distances (primary)
d_osrm <- surveyzones_compute_sparse_distances(
  access_points = access_points,
  engine = surveyzones_engine_osrm(measure = "duration"),
  chunk_size = 50
)

# Haversine distances (fallback)
d_haversine <- surveyzones_compute_sparse_distances(
  access_points = access_points,
  engine = surveyzones_engine_haversine()
)

# Combine: OSRM preferred, haversine * 60 as fallback
d_combined <- d_haversine |>
  left_join(d_osrm,
            by = join_by(origin_id, destination_id),
            suffix = c("_haversine", "_osrm")) |>
  mutate(distance = coalesce(distance_osrm, distance_haversine * 60))
```

### Zone construction and sequencing

Each state is solved independently via `enforce_partition = TRUE`. The
`uncap_then_split` strategy first finds the minimum number of travel groups
per state, then splits oversized groups to respect the 6-UPA limit.

```{r}
#| label: pns-zones
#| eval: false

plan <- surveyzones_build_zones(
  sparse_distances      = d_combined,
  tracts                = tracts,
  D_max                 = 10,
  max_workload_per_zone = 6,
  enforce_partition      = TRUE,
  solver                = "highs",
  max_time              = 180,
  rel_tol               = 0.01
)

plan <- surveyzones_sequence(plan, d_combined)
```

### Results

```{r}
#| label: pns-results
#| echo: false
#| eval: !expr requireNamespace("pns.zonas", quietly = TRUE)

library(pns.zonas)
library(surveyzones)
library(dplyr)

plan_full <- readr::read_rds(file.path(rdir, "vignettes/zonas.rds"))

uf_siglas <- c(
  "11" = "RO", "12" = "AC", "13" = "AM", "14" = "RR", "15" = "PA",
  "16" = "AP", "17" = "TO", "21" = "MA", "22" = "PI", "23" = "CE",
  "24" = "RN", "25" = "PB", "26" = "PE", "27" = "AL", "28" = "SE",
  "29" = "BA", "31" = "MG", "32" = "ES", "33" = "RJ", "35" = "SP",
  "41" = "PR", "42" = "SC", "43" = "RS", "50" = "MS", "51" = "MT",
  "52" = "GO", "53" = "DF"
)

resumo <- plan_full$zones |>
  mutate(uf_codigo = substr(zone_id, 1, 2)) |>
  summarise(
    n_zonas = n(),
    upas_total = sum(n_tracts),
    tamanho_mediano = median(n_tracts),
    diametro_mediano = median(diameter, na.rm = TRUE),
    n_zonas_1upa = sum(n_tracts == 1),
    pct_zonas_1upa = 100 * mean(n_tracts == 1),
    .by = uf_codigo
  ) |>
  mutate(UF = uf_siglas[uf_codigo]) |>
  arrange(uf_codigo) |>
  select(
    UF, Zones = n_zonas, UPAs = upas_total,
    `Median size` = tamanho_mediano, `Median diam. (min)` = diametro_mediano,
    `1-UPA zones` = n_zonas_1upa, `% 1-UPA` = pct_zonas_1upa
  )

totais <- tibble::tibble(
  UF = "Total",
  Zones = sum(resumo$Zones),
  UPAs = sum(resumo$UPAs),
  `Median size` = median(plan_full$zones$n_tracts),
  `Median diam. (min)` = median(plan_full$zones$diameter, na.rm = TRUE),
  `1-UPA zones` = sum(resumo$`1-UPA zones`),
  `% 1-UPA` = round(100 * sum(resumo$`1-UPA zones`) / sum(resumo$Zones), 1)
)

resumo <- bind_rows(resumo, totais)

tinytable::tt(resumo,
  caption = "Zone summary by state for the PNS biomarker collection.",
  digits = 1
) |>
  tinytable::style_tt(i = nrow(resumo), bold = TRUE)
```

The framework partitions the 3,004 UPAs across 27 states into 905 zones that
respect both the 10-minute radius bound ($D_{\max}$) and the 6-UPA workload
limit. States
with dense urban cores (e.g., DF, RJ) produce larger zones with lower
diameters; states with dispersed metropolitan regions (e.g., AM, PA) produce
more single-UPA zones where geographic isolation makes grouping infeasible.

All 27 state-level sub-problems reached optimality. Total solve time across
all states was 1,269 seconds (median 36 seconds per state, range 8--147
seconds). The most expensive sub-problem was Espírito Santo (state 32, 147
seconds). These timings include the full `uncap_then_split` pipeline
(Phase 1 $K$-search + Phase 2 splitting) using HiGHS on a single core.



# Discussion {#sec-discussion}

## Contributions

The contribution of surveyzones is not a new facility location formulation but an operational framework that combines radius-constrained assignment, capacity constraints, and automatic $K$-determination within a decomposition architecture tailored to survey zoning problems. The `uncap_then_split` decomposition (@sec-k-discovery) converts a large capacitated location problem into a sequence of smaller tractable sub-problems while producing a travel group hierarchy that supports three-level sequencing and human-readable zone identifiers.

## Partitioning as an operational tool

The `enforce_partition` mechanism is flexible beyond the two applications
demonstrated here: it can encode any categorical constraint---agency
jurisdiction (ensuring that UPAs from distinct agencies are never grouped
together), area type (solving favelas separately from regular urban areas,
given substantially different access and travel time profiles), or
administrative boundaries---without changes to the solver pipeline.

## Limitations and possible extensions

The current framework assumes that workload per tract is known in advance and
static. In adaptive survey designs where workload depends on response rates
observed during data collection, dynamic re-zoning may be needed. The
framework's caching mechanism could support incremental re-optimization, but
this is not yet implemented.

The $D_{\max}$ pre-filtering is a hard constraint: tracts beyond $D_{\max}$
from all candidate centers cannot be assigned and are flagged as isolated
components. In practice these are rare (islands, remote areas) and are handled
as single-tract zones, but a soft-constraint formulation could provide more
flexibility.

The MILP formulation does not enforce zone contiguity explicitly; contiguity
is achieved indirectly through $D_{\max}$ pre-filtering and the
sum-of-distances objective, which naturally favors compact, connected zones.
For applications where non-contiguous zones are unacceptable, explicit
contiguity constraints could be added at the cost of increased model
complexity.

The `uncap_then_split` heuristic trades global optimality for computational
tractability and the travel group hierarchy. The optimality gap---the
difference between the zone count produced by `uncap_then_split` and the
true minimum under both constraints---is zero when no Phase 1 group exceeds
$W_{\max}$ (no splitting needed), and grows with the number of oversized
groups. In the PNS application, where workload per UPA is relatively uniform
and $W_{\max}$ is generous (6 UPAs), most groups fit within capacity
directly; the gap is driven by a small number of dense urban clusters.
The `direct` strategy (@sec-direct) provides a comparator for the minimum
zone count under the same constraints and can be used to quantify this gap on
a case-by-case basis
though at higher computational cost for large instances.


# Conclusion {#sec-conclusion}

We presented surveyzones, an integrated framework for constructing
workload and geographically bounded survey field zones. The framework
combines a capacitated p-median MILP with $D_{\max}$ pre-filtering, a
two-phase decomposition that decouples geographic feasibility from workload
balancing, and a three-level TSP-based sequencing algorithm. We demonstrated
the framework on two applications: census tract zoning for a household survey
in Rio de Janeiro and biomarker collection logistics for the Brazilian
National Health Survey across 27 state capitals.

The framework is implemented as an open-source R package (`surveyzones`). It
supports pluggable distance engines (OSRM,
haversine), multiple MILP solvers (HiGHS, GLPK, CBC), disk caching for
reproducibility, and flexible partitioning by arbitrary administrative or
operational criteria.


# References {.unnumbered}

::: {#refs}
:::
